---
title: "Broadway Shows: Earnings, Attendance and Show Statistics (1995 - 2015)"
subtitle: "CS544 Spring 1 Online Term Project"
author: "Lauren Savrin"
date: "February 29th, 2024"

output:
  html_document:
    fig_width: 8
    code_folding: hide
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
  pdf_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4,
                      echo=FALSE, warning=FALSE, message=FALSE)

library(plotly)
library(sampling)
library(dplyr)

broadway<-read.csv("/Users/lsav1/OneDrive/Desktop/broadway.csv")
twenty<-subset(broadway,broadway$Date.Year>=1995 & broadway$Date.Year<=2015)
```


# Data Set Overview and Prep

The data set used for this analysis was pulled from Kaggle (Broadway Shows) and captures Broadway show data from 1990 to 2016. It includes data that captures the show attributes like the date, name, type and theater. Additionally, the set includes statistics of the weekly attendance, earnings, capacity and the number of performances that occurred each week. The show information is broken out by week which creates some potential for duplication and there needs to be some manipulation to ensure that the data extracted and used for analysis is accurate. In order to ensure that the data was consistent and manageable, the data set was trimmed down to only include the information from 1995 to 2015 by establishing a subset the falls within those years. This was done because the data in the early years of the overall set was sparse and likely lacking information. The 2016 data was only available through August of that year so it was removed to be able to capture accurate annual data. The analysis was then done on this 20 year subset of the overall data set.

# Objective

The objective of this project is to gain further knowledge on Broadway trends and earnings.


# Show Type Distribution
One of the areas that peaked my interest is the distribution of shows across the 3 primary show types: Plays, Musicals and Specials. In order to ensure that the information is accurate, for this analysis we needed to eliminate double counting the type for a singular show across a several week run. In order to do so, a modified set was generated by counting the show type and then the show name to ensure that the count only captures each show once. From there we capture the division based on show type and then how that division is affected when the total earning from shows is taken into account.

## 20 Year Distribution Based on Volumes

The following chart captures the volume of each category of show in the period outlined. With so many shows that come across the Great White Way, I chose this variable because I wanted to understand the distribution of the different types. As a theater fan who leans more towards musicals than other stage performances, I was curious to see what winds up on stage. In order to understand this distribution, we needed to adjust the data set to ensure that the types are not double counted for individual shows. The best way to represent the output for this division of information is in a pie chart as each show is attributed a single type. The following chart shows the percentage of all shows that are attributed to each type.  

```{r echo=TRUE}
show_type<-twenty|>
  count(Show.Type,Show.Name)
Type<-as.data.frame(table(show_type$Show.Type))
plot_ly(labels = ~Type$Var1, values = ~Type$Freq,
              marker=list(colors=c("cadetblue","darkorchid","coral")),
              title= "Type of Show",type="pie")
```

The majority of the shows in this 20 year period are plays at 56.4% of the productions staged. Musicals make up 37.4% of all shows. Musicals are significantly more complicated to get prepared for and more costly to stage. Plays require actors, a theater and a script to move from concept to stage. Musicals require singing, dancing and acting so more staff are needed to get things up and running and to actual put the show on 8 times per week. The last category that makes up less than 10% of the productions staged are specials. As their name indicates, the minimal frequency of these production types is unsurprising because they tend to be limited runs. 

## 20 Year Distribution by Total Gross

The below chart takes the gross earnings per show type by grouping the data set by show type and then summing the gross earnings for each show. The data is then loaded into a pie chart based on the show type to capture the difference from the division by count. In order to capture the total spend per type, the sum needed to be divided by 1,000 to avoid numbers that were too large to properly present. While plays make up the majority of productions staged in this 20 year period, I wanted to understand if that held true when the gross for the shows was taken into account. 

```{r echo = TRUE}
By_Show_Spend<-twenty |>
  group_by(Show.Type) |>
  summarize(total_spend_show=sum(Statistics.Gross))
plot_ly(By_Show_Spend,labels = By_Show_Spend$Show.Type, 
               values = ~By_Show_Spend$total_spend_show/1000,
               marker=list(colors=c("cadetblue","darkorchid","coral")),
               title = "Total Spend by Show (in thousands)", type="pie")
```

As expected the portion attributed to each show type changes significantly when the total earnings is taken into account. Musicals make up 84.4% of the total earnings and becomes the majority component in the division of this breakdown. Play earnings make up only 14.9% of total earnings and specials make up less than 1% of the total earnings. It is interesting that more plays are staged but are earning significantly less than musicals. Musicals are more costly to stage than plays and, as a result, the musical tickets are often more expensive than play tickets. Furthermore, depending on the play and the theater, the number of seats available is likely lower for a theater that produces plays. Musicals may have more weekly productions and can have longer runs while drawing more patrons from New York City and elsewhere. 

# Evaluation of Gross Earnings

The next component that was evaluated was to dive deeper into the earnings breakdown by show type and by year to see how the total earnings changed year over year and how the division by show type is impacted by each year's productions.

## Annual Show Type Distribution

The yearly total earnings increases as time goes on with a general upward trend. While a few years saw dips in earnings as time went on the trend is regularly upward. The years that differ from this increase trend are 2001 and 2009. With 9/11 occurring in 2001, the ripple effect from that event limited travel to New York City and therefore patrons for Broadway productions. 2009 was a period of recovery from the Great Recession, as people started to get back on their feet, they were more cautious with how they spen their money. Throughout all of the years, the division by show type aligns with the pie chart from above with musicals making up the largest portion of earnings. The earnings by Musicals and Plays also followed increasing trends per year with 2014 and 2015 containing the highest earnings for those two show types.

```{r echo = TRUE}
By_Show<-twenty |>
  group_by(Date.Year,Show.Type) |>
  summarize(total_spend_show=sum(Statistics.Gross))
plot_ly(By_Show,x = ~Date.Year, y = ~total_spend_show, type="bar",color = ~factor(Show.Type))|>
  layout(barmode="stack", title = "Total Earnings by Show Type and Year", xaxis=list(title="Year"), yaxis = list(title="Total Spend ($)"))

```

## Gross Distribution by Month and Year

In addition to monitoring earnings per year and show type, I wanted to take that annual information and review it based on the earnings per month. While the earnings vary per year the general trend is that earnings are higher in the beginning and end of the year than in the middle of the year. January earnings are larger as shows that had tickets purchased as gifts for the holidays are often running in January. November and December earnings are higher than the rest of the year. With the holiday season there are the regular shows, the holiday specials and more people traveling to visit family, celebrate the holiday season, or enjoy their vacations. The July , August and September earnings are lower across the years as people are more likely to spend their time outdoors, on seasonal vacations and many families have children that are off at summer camps. The number of family trips and likelihood for attendance and travel to New York City to go to the theater is likely lower in the summertime. Even with all of those trends there is never a month across all of those years where Broadway's earnings were not in the multi-million dollar range.

```{r echo = TRUE}
Yearly<-twenty |>
  group_by(Date.Year,Date.Month) |>
  summarize(total_spend=sum(Statistics.Gross))
plot_ly(Yearly,x = ~Date.Year, y = ~total_spend, type="bar",color = ~factor(Date.Month))|>
                 layout(barmode="stack", title = "Total Earnings by Month and Year", xaxis=list(title="Year"), yaxis = list(title="Total Spend ($)"))

```




# Top 5 Shows by Earnings (Chicago, Mamma Mia, The Lion King, Phantom of the Opera and Wicked)

## Weekly Earnings Distributions

There are 795 shows included in this set. While it would be great to look at the weekly earnings distribution for all of the shows that would not be easy to capture visually. As a result I decided to determine the 5 shows in this 20 year period that earned the most to evaluate the earnings distribution. The 5 shows that earned the most are not surprising as they are some of the longest running shows on Broadway and either ran for most of this time frame or had multiple runs come back through Broadway. The earnings information that was ultimately evaluated were the weekly earnings for each of those 5 shows across the 20 year period. The Earnings information varies tremendously even among these top earning shows. 

```{r echo = TRUE}
Gross<- twenty |>
  group_by(Show.Name) |>
  summarize(Earnings=sum(Statistics.Gross)) |>
  arrange(desc(Earnings))
top5<- Gross[1:5,]
top5_detail<-subset(twenty, twenty$Show.Name %in% top5$Show.Name)

plot_ly(top5_detail, y = ~Statistics.Gross, type = "box", color = ~Show.Name) %>%
  layout(title = "Earnings for Top 5 Shows", yaxis = list(title = "Total Weekly Earnings"))

```

## Findings
 1. All of the shows have a series of outliers for their weekly earnings which is unsurprising based on the uncertain nature of Broadway productions, theater capacity and attendance based on people in the city that are looking to see shows. 
 2. There is a resource that sells same day tickets for a discount for people that do not have a strong preference of show and are willing to make a last minute decision that could impact the weekly earnings. These 5 shows often have tickets available at this resource and while they undoubtedly sell tickets in advance for all of the shows there will always be weeks where the show is more popular than others. 
 3. The earnings for Chicago is the most compacted set of data which shows that the weekly earnings for this show is more consistent than others in this subset. Chicago's distribution is the most even from the median with the Q1 and Q3 at about $75k from the median and the fences at about $280k from the median.
 4. Mamma Mia and Phantom are both long running shows that are very popular but do usually appeal to a portion of the population. Phantom is a classic with one of the longest Broadway runs in history. With a show like that where there is consistent attendance for so many years, the earnings are expected to be a bit more consistent. It has a significant number of outliers on the high end of the range where the show had particularly lucrative weeks.
 5. Mamma Mia! has one of the densest set of outliers on both the top and bottom edges of the range. This indicates that the weekly earnings is much more likely to fluctuate than some of the other shows.
 6. Some of these shows, especially when they have run for as long as Phantom has, tend to attract celebrity talent which therefore attracts more attendees and can lead to inflated tickets prices and, as a result, higher weekly earnings. 
 7. Lion King earnings is the largest range which is interesting because it was a one of a kind production when it was originally staged. The production involves intricate puppetry which differentiated it from actors simply playing a character rather than having to embody a wild animal. The Lion King opened on Broadway in 1997 and appealed to a wider range of patrons. As more and more people realized what a fantastic show it was and it gained more traction, the tickets could get more expensive and the producers could be confident that people would still attend. The first to third quartile is the widest range of all of these shows because the weekly earnings have been impacted by the continued run through most of this data set and the growth in popularity. 
 8. Wicked was another show that gained popularity with speed and drew audiences from all over the world. It can be attended by families and individuals alike which draws a broader and therefore larger audience. Wicked premiered on Broadway in 2003 and was nominated for best musical in 2004. For those that follow the Tony Awards, musicals that receive even just a nomination often see an influx of sales following the awards. The connection to the classic Wizard of Oz story does not hurt either. Wicked had the top earning week at $3.2M but a total range of $821k to $2.2M with a median at $1.5M which makes it relatively evenly distributed within that majority range around the median. It also includes the broadest range of outliers, which further cements that popularity and traction can influence the overall spend per show.


## Top 5 Sampling
With this Top 5 data, I decided to evaluate this using a series of sampling methods to see if what was extracted aligned with the larger data set as it was distributed across these 5 shows. The sampling methods used here are simple random sampling without replacement, systematic, and stratified sampling. The total pool includes the following breakdown: Chicago at 999, Mamma Mia! at 725, The Lion King at 947, The Phantom Of The Opera at 1,020 and Wicked at 636. The sample size of 500 is being used for each sampling method.

Simple random sampling without replacement pulls a specified sample set from the larger set with each element having the same opportunity to get selected from the larger data set. 500 instances will be randomly selected without replacement.

The next method of sampling is systematic sampling. Systematic sampling is when there is a set sampling interval that guides the sample selections. The likelihood of bias with this type of sampling is higher than some other sampling methods.

The final method used is stratified sampling. Stratified sampling is when the data set is broken into smaller groups and then certain sizes are picked from each group to keep the distribution across the categories in line with the larger pool.

From left to right and then top to bottom, the plots represent the densities compared to the entire population, then simple random sampling without replacement, systematic sampling, and finally, stratified sampling.


```{r echo = TRUE}
sampling<- top5_detail
sampling_den<-table(sampling$Show.Name)/nrow(sampling)


m1<- mean(sampling$Statistics.Attendance)
s1<-sd(sampling$Statistics.Attendance)

paste("Population:",names(sampling_den),"represents", round(sampling_den/sum(sampling_den)*100),"%","of the set",sep=" ")

plot11<-plot_ly(sampling, x=names(sampling_den),y=sampling_den, type="bar", name="Population")

set.seed(1931)
sample.size<-500
srs <- srswor(sample.size, nrow(sampling))
rows <- (1:nrow(sampling))[srs!=0]
sample.1 <- sampling[rows, ]

f<-table(sample.1$Show.Name)
densities<-f/sample.size
paste("SRSWOR:",names(f),"represents", round(f/sum(f)*100),"%","of the set",sep=" ")

srsm<-mean(sample.1$Statistics.Attendance)
srssd<-(sd(sample.1$Statistics.Attendance))

plot12<-plot_ly(sample.1, x=names(densities),y=densities, type="bar", name="SRSWOR")

#Systematic Sampling
set.seed(1931)
pik <- inclusionprobabilities(sampling$Statistics.Attendance,sample.size)
set.seed(1931)
d<-(UPsystematic(pik))
sample.2 <- sampling[d != 0, ]

freq<-table(sample.2$Show.Name)
den<-freq/sample.size

paste("Systematic Sampling:", names(freq),"represents",round(freq/sum(freq)*100),"% of the set",sep=" ")

plot13<-plot_ly(sample.2, x=names(den),y=den, type="bar", name="Systematic Sampling")


ssm<-mean(sample.2$Statistics.Attendance)
sssd<-(sd(sample.2$Statistics.Attendance))


#Stratified Sampling
order.index <- order(sampling$Show.Name)
data <- sampling[order.index, ]
t<-table(data$Show.Name)
set.seed(1931)
stratum_sample_sizes <- round(sample.size * (t / sum(t)))

set.seed(1931)


sample.3 <- sampling::getdata(data, (strata(data, stratanames = c("Show.Name"),
                     size = stratum_sample_sizes, method = "srswor",
                     description = FALSE)))

freq1<-stratum_sample_sizes/sum(stratum_sample_sizes)
paste("Stratified Sampling:", names(freq1),"represents",round(freq1*100),"% of the set",sep=" ")

plot14<-plot_ly(sample.3, x=names(freq1),y=freq1, type="bar", name="Stratified Sampling")


stm<-mean(sample.3$Statistics.Attendance)
stsd<-sd(sample.3$Statistics.Attendance)

subplot(plot11, plot12, plot13, plot14, nrows = 2, titleX = TRUE)|>
  layout(title="Densities by Different Sampling Types on the Top 5 Shows", xaxis=list(tickangle=-12),xaxis2=list(tickangle=-12),xaxis3=list(tickangle=-12),xaxis4=list(tickangle=-12))

samp_means<-c(m1,s1,srsm,srssd,ssm,sssd,stm,stsd)
samp_name<-c("Population","","SRSWOR","","Systematic Sampling","","Stratified Sampling")
samp_steps<-seq(1,length(samp_means),by=2)
for (i in samp_steps){
  cat(samp_name[i],":","The mean is",samp_means[i],"and the standard deviation is",samp_means[i+1],"\n")}


```
When looking at the different sampling methods I also compared means and standard deviations based on the attendance rates for each of the weeks. The attendance means for the samples were close to the mean for the population with the systematic sampling being the farthest off. This aligns with the way the sampling densities for that set vary from the population. The standard deviations are all relatively consistent as well which simply means that the distribution range for weekly attendance is close to the same general range across the different sampling methods.


# Average Ticket Pricing - Central Limit Theorem
In order to evaluate the Central Limit Theorem I generated a new variable called Average Ticket Price which takes the weekly gross and divides it by the weekly attendance for that show. Obviously different shows have different ticket prices and different seats within those shows have different prices but, since we are unable to determine the individual seat pricing, the closest we can get is determining the average ticket price. This information is captured in the histogram below which shows that the average ticket price is skewed to the right which means that most of the ticket pricing falls on the lower half of the overall range and it tapers off as the average ticket prices gets higher. This distribution is expected because there are more seats available at a lower price points to make theater accessible to all. Additionally, we have to take into account that this covers the period of time between 1995 and 2015 and ticket prices associated with Broadway production continued to increase each year. The mean indicator is included in the following plot and represented by the black line and falls between $70 and $71.

```{r echo = TRUE}
b<- twenty |>
  mutate(avg_ticket_price=Statistics.Gross/Statistics.Attendance)
m<-mean(b$avg_ticket_price)
s<-sd(b$avg_ticket_price)
plot_ly(b, x=b$avg_ticket_price, type="histogram")|>
  add_lines(y = range(c(0,700)),x = mean(b$avg_ticket_price),line = list(color = "black"),inherit = FALSE, showlegend = FALSE)|>
  layout(title="Average Ticket Price", yaxis=list(title="Volumes"), xaxis=list(title="Average Ticket Price"))


```

## CLT Test
The below captures samples of the average ticket prices for 4 different samples sizes: 100, 300, 500 and 700. As the sample size gets larger the distribution of ticket prices gets more concentrated around the mean value of the population data set which is indicated by the vertical black lines in the graphs below. With different sample sizes it is important to capture a range but ensure that the graphs are being viewed on the same axis distributions to allow for accurate comparisons. The mean of each sample set aligns very closely to the mean of the larger set. The standard deviation is largest in the overall data set. When looking at the different sample sizes, the larger sample sizes align to smaller standard deviations. This indicates that, as the sample sizes increase, the data is getting more concentrated around the mean line. By definition most data should fit within 3 standard deviations from the mean, so with a smaller standard deviation, the majority of the data falls in a smaller range.
```{r echo = TRUE}
set.seed(1931)
samples<-1000
sample.size1<-100
sample.size2<-300
sample.size3<-500
sample.size4<-750
bar1 <- numeric(samples)
bar2 <- numeric(samples)
bar3 <- numeric(samples)
bar4 <- numeric(samples)

for (i in 1: samples) {
  bar1[i] <- mean(sample(b$avg_ticket_price, sample.size1, replace = FALSE))
  bar2[i] <- mean(sample(b$avg_ticket_price, sample.size2, replace = FALSE))
  bar3[i] <- mean(sample(b$avg_ticket_price, sample.size3, replace = FALSE))
  bar4[i] <- mean(sample(b$avg_ticket_price, sample.size4, replace = FALSE))
}

df1<-as.data.frame(bar1)
df2<-as.data.frame(bar2)
df3<-as.data.frame(bar3)
df4<-as.data.frame(bar4)

plot7<-plot_ly(df1, x=df1$bar1, type="histogram", name= "Sample Size 100") |>
  add_lines(y = range(c(0,100)),x = mean(b$avg_ticket_price),line = list(color = "black"),inherit = FALSE, showlegend = FALSE) |>
  layout(yaxis = list(tickvals = c(0,25,50,75,100), ticktext = c(0,25,50,75,100), range = c(0, 100)),
         xaxis = list(tickvals = c(60,65,70,75,80), ticktext = c(60,65,70,75,80), range = c(60, 80)))
plot8<-plot_ly(df2, x=df2$bar2, type="histogram",  name= "Sample Size 300") |>
  add_lines(y = range(c(0,100)),x = mean(b$avg_ticket_price),line = list(color = "black"),inherit = FALSE, showlegend = FALSE) |>
  layout(yaxis = list(tickvals = c(0,25,50,75,100), ticktext = c(0,25,50,75,100), range = c(0, 100)),
         xaxis = list(tickvals = c(60,65,70,75,80), ticktext = c(60,65,70,75,80), range = c(60, 80)))
plot9<-plot_ly(df3, x=df3$bar3, type="histogram", name= "Sample Size 500")|>
  add_lines(y = range(c(0,100)),x = mean(b$avg_ticket_price),line = list(color = "black"),inherit = FALSE, showlegend = FALSE) |>
  layout(yaxis = list(tickvals = c(0,25,50,75,100), ticktext = c(0,25,50,75,100), range = c(0, 100)),
         xaxis = list(tickvals = c(60,65,70,75,80), ticktext = c(60,65,70,75,80), range = c(60, 80)))
plot10<-plot_ly(df4, x=df4$bar4, type="histogram", name= "Sample Size 700")|>
  add_lines(y = range(c(0,100)),x = mean(b$avg_ticket_price),line = list(color = "black"),inherit = FALSE, showlegend = FALSE) |>
  layout(yaxis = list(tickvals = c(0,25,50,75,100), ticktext = c(0,25,50,75,100), range = c(0, 100)),
         xaxis = list(tickvals = c(60,65,70,75,80), ticktext = c(60,65,70,75,80), range = c(60, 80)))

subplot(plot7, plot8, plot9, plot10, nrows = 2, titleX = TRUE) |>
  layout(title="Average Ticket Price for Different Sample Sizes")

```


```{r echo = TRUE}
set.seed(1931)

cm<-mean(b$avg_ticket_price)
cs<-sd(b$avg_ticket_price)
cm1<-mean(bar1)
cm2<-mean(bar2)
cm3<-mean(bar3)
cm4<-mean(bar4)
cs1<-sd(bar1)
cs2<-sd(bar2)
cs3<-sd(bar3)
cs4<-sd(bar4)

means<-c(cm,cs,cm1,cs1,cm2,cs2,cm3,cs3,cm4,cs4)
Name<-c("Original","","Sample 100","","Sample 300","","Sample 500","","Sample 700")
steps<-seq(1,length(means),by=2)
for (i in steps){
  cat(Name[i],":","The mean is",means[i],"and the standard deviation is",means[i+1],"\n")}

```

# Average Gross Per Show

This last scatter plot captures the average earning for all of the shows in the data set. The distribution of this data is broad but I have included a line that outlines the mean of all of the means at $390k. Based on the placement of this line, it is clear that the concentration of means below it is significantly denser than those above it because there is more space to occupy above that mean.

One might expect that the shows that wind up at the top would be similar to the top 5 by total earnings but actually the Top 5 shows here are Aladdin, Hamilton, Hugh Jackman: Back on Broadway, The Book of Mormon, and Wicked. Wicked is the only overlap between these two 'top 5' segments. This data is extrapolated by calculated the mean of the Weekly Gross for each show title. With 795 shows in the data set they cannot all be major hits that bring in a lot each week. There are many shows on Broadway, likely those that are lowest in this chart that had limited runs or where closed early due to limited engagement. 

```{r echo = TRUE}
Total<-twenty |>
  group_by(Show.Name) |>
  summarize(mean=mean(Statistics.Gross))
plot_ly(Total,x = ~Total$Show.Name, y = ~Total$mean, type="scatter", mode="markers") |>
  layout(title="Average Weekly Show Earnings", xaxis = list(title = "", showticklabels = FALSE), yaxis=list(title="Average Gross by Show"))|>
  add_lines(x = Total$Show.Name,y = mean(Total$mean),line = list(color = "darkgrey"),inherit = FALSE, showlegend = FALSE)


```

# Conclusions

This was a relatively large data set and it was fascinating to break the data down and see how the different values are impacted by each other. I have highlighted some of the standout findings below.

  1. The most popular and longest running shows were the top earners. 
  2. While musicals are not the largest volume of productions in this 20 year periods, they were the largest earners.
  3. The average ticket pricing meets the Central Limit Theorem for the sample sizes and aligns around a consistent mean.
  4. The total earnings across all shows increased over time and aligned to the distributions that we expected based on the show type ratio and the traffic across the different months.
  5. When looking at pulling samples from the Top 5 data set that was generated, the samplings aligned relatively closely with the population distribution.

With so many factors that can impact the success or failure of a production from a functional perspective it is interesting to focus on the positive outcomes and evaluate how this data can be segmented and reviewed to determine different areas of impact.



